Quality Definitions of Qualities

### 2021-09-20 StateOfPractice/QDefOfQualities/QDefOfQualities.tex
Updates to quality of qualities paper, reformat, start introduction

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}

\usepackage{hyperref}
\hypersetup{
colorlinks=true,        % false: boxed links; true: colored links
linkcolor=red,          % color of internal links (change box color with linkbordercolor)
citecolor=blue,         % color of links to bibliography
filecolor=magenta,      % color of file links
urlcolor=cyan           % color of external links
}

@@

@@

\begin{abstract}
	...
\end{abstract}

@@

Purpose and scope of the document.

\wss{ Needs to be filled in.  Should reference the overall research proposal,
	and the ``state of the practice'' exercise in particular.}

Given that our research focuses on improving the quality of software, we often
find it necessary to have clear definitions of the different software qualities,
like maintainability, usability, sustainability, etc. Unfortunately, the
definitions that we find in the literature are not always satisfying.  They are
not always adequate for unambiguous communication, or they simply are	not useful
for providing guidance on what software decisions will lead to quality
improvement. Moreover, methods for measuring qualities based on the definitions
are not always clear.  Therefore, we have endeavored to build a lis of
definitions that can aid and guide our research. We are publishing our
collection, since others may likewise find them valuable.

Methodology:
\begin{itemize}
  \item Collect the list of qualities to define, based on the qualities that are
  listed in software engineering texts, such as Ghezzi Et al and VanHliet.
  \item For each quality find as many definitions as possible
  \item Based on the existing definitions, propose a high quality definition of
  each quality.  The proposed definition may use one of the existing, or combine
  several of the options.  We may also make other modifications.  The goal is
  for the proposed definition to address our acceptance criteria.
\end{itemize}

@@

@@

\begin{description}
  \item[Unambiguous]
  \item[Consistency] The final list of definitions should be consistent in the
  terminology used.  For instance, we do not use synonyms for variety; we use
  the same word to mean the same concept throughout the final list of
  definitions.  This is why measurement is given by ``the effort required..''
  and not the various synonyms for this that are seen in the literature. \wss{We
  may change this phrase we use for measuring something.  I just wanted to put
  this in here as a placeholder.}
  \item[Generality]
  \item[Measurability]
  \item[Etc.]
\end{description}

@@

@@

\subsection{{Installability} \oo{owner}}

\begin{defn}
  The capability of the software product to be installed in a specified
  environment \citep{ISO9126}.
\end{defn}

\begin{defn} \label{Installability_Selected2}
	The degree of effectiveness and efficiency with which a product or system can
	be successfully installed and/or uninstalled in a specified environment
	\citep{ISO/IEC25010}.
\end{defn}

@@

@@

\subsection{{Correctness} \oo{owner}}

@@

@@

\begin{defn} \label{Defn_Verifiability1}
Verifiability ``means that you should be able to write a set of tests that can
demonstrate that the delivered system meets each specified requirement'' \citep{sommerville}.	
\end{defn}

@@

\begin{defn}
	Verifiability refers to how well software components or the integrated
	product can be evaluated to demonstrate whether the system functions as
	expected \citep{wiegers2003softreq}.
\end{defn}
\noindent \textbf{Proposed Definition}
Definition~\ref{Defn_Verifiability1} rephrased:
Verifiability is the extent to which a set of tests can be written and executed, to demonstrate
that the delivered system meets the specification.

@@

Definition is concise and measurable.  Verifiability involves solving the
equations right \citep[p.~23]{Roache1998} and the definition states that system needs to be verifiable based on the specifications..

@@

\begin{defn}
  The assurance that a product, service, or result meets the needs of the customer and other identified
  stakeholders \citep{project2017guide}.
\end{defn}
\begin{defn}
	Software validation is the process of checking that the system conforms to its specification and
	that it meets the real needs of the users of the system \citep{sommerville}.

\end{defn}
\begin{defn}
	Software validation, where the software is checked to ensure that it is what the
	customer requires \citep{sommerville}.
\end{defn}
\begin{defn}
  Software validation is achieved through a series of tests that demonstrate conformity
  with requirements \citep{pressman2005software}.
\end{defn}

@@

@@

Definition describes that system should be measurable through the development life cycle,
It is concise and can be measured.
\subsection{\notdone{Testability} \oo{owner}}

@@

Definition is concise and measurable

@@

@@

\begin{defn}
 The capability of the software product to maintain a specified level of performance when used under specified conditions \cite{ISO9126}.
\end{defn}
\begin{defn} \label{reliabilitySelected1} It is the probability of failure-free
  operation of a computer program in a specified environment for a specified
  time \citep{musa1987software}.
\end{defn}

@@

\begin{defn}
  Informally, the reliability of a system is the probability, over a given
  period of time, that the system will correctly deliver services as expected by
  the user \citep{sommerville}.
\end{defn}

\begin{defn}
  The ability of a system to perform its required functions under stated
  conditions whenever required—having a long mean time between failures
  \citep{mcconnell2004code}.
\end{defn}
\begin{defn} \label{reliabilitySelected2} Reliability is concerned with measuring
	the probability of occurence of faliure - that is, the probability of the
	observable effects of errors \citep{GhezziEtAl2003}.
\end{defn}
\oo{Statistical definition from  \citep{GhezziEtAl2003} added}

@@

\subsection{{Robustness} \pmi{owner}}
\begin{defn}
	The degree to which a system or component can function correctly in the
	presence of invalid inputs or stressful environmental conditions
	\citep{IEEEStdGlossarySET1990}.
\end{defn}
\begin{defn}
	The quality can be further informally refined as the ability of a software to
	keep an acceptable behaviour, expressed in terms of robustness requirements,
	in
	spite of exceptional or unforeseen execution conditions (such as the
	unavailability of system resources, communication failures, invalid or
	stressful inputs, etc.) \citep{fernandez2005model}.
\end{defn}
\begin{defn} \label{RobustnessPartialDefn}
	Code possesses the characteristic of robustness to the extent that it can
	continue to perform despite some violation of the assumptions in its
	specification \citep{boehm2007software}.
\end{defn}
\begin{defn} \label{RobustnessDefnSelected}
	A program is robust if it behaves ``reasonably'', even in circumstances that
	were not anticipated in the requirements specification - for example, when it
	encounters incorrect input data or some hardware malfunction
	\citep{ghezzi1991fundamentals}.
\end{defn}

@@

\begin{defn} \label{PerformanceDefnSelected}
	The degree to which a system or component accomplishes its designated functions
	within given constraints, such as speed, accuracy, or memory usage
	\citep{IEEEStdGlossarySET1990}.
\end{defn}
\begin{defn} \label{PerformanceDefnSelected2}
	How well or how rapidly the system must perform specific functions. Performance
	requirements encompass speed (database response times, for instance), throughput
	(transactions per second), capacity (concurrent usage loads), and timing (hard
	real-time demands) \citep{wiegers2003softreq}.
\end{defn}
\begin{defn}
	In software engineering we often equate performance with efficiency. A software system is efficient if it uses computing resources economically
	\citep{ghezzi1991fundamentals}.
\end{defn}

@@

\pmi{I will check if Wiegers discusses the difference between speed and timing.}
\pmi{Wiegers did not discuss this difference.}
\pmi{Speed - change over time; timing - related to specification: when something
	should happen in relation to something else - state transitions}
\wss{If we need both speed and timing, we should define them.  We aren't
  including accuracy from the IEEE definition.  Is that on purpose?}
\pmi{We have decided not to include accuracy. SS will find difference b/w speed and timing from a colleague.}

@@

@@

Nielsen and (separately) Schneidermann have defined usability as part of
usefulness and is composed of:
\begin{defn}
\begin{itemize}
\item Learnability: How easy is it for users to accomplish basic tasks the first
  time they encounter the design?
\item Efficiency: Once users have learned the design, how quickly can they
  perform tasks?
\item Memorability: When users return to the design after a period of not using
  it, how easily can they re-establish proficiency?
\item Errors: How many errors do users make, how severe are these errors, and
  how easily can they recover from the errors?
\item Satisfaction: How pleasant is it to use the design?
\end{itemize}
In that context, it makes sense to separate \emph{usefulness} into
\emph{usability} (purely an interface concern) and \emph{utility} (in the
economics sense of the word).
\cite{NielsenNorman}
\end{defn}

@@

\begin{defn}
  ``A software is usable - or user friendly - if its human users find it easy
  to use'' \citep{ghezzi1991fundamentals}.
\end{defn}

@@

@@

\begin{defn}
  ISO/IEC 25010 refers to maintainability as the degree of effectiveness and
  efficiency with which a product or system can be modified by the intended
  maintainers \citep{ISO/IEC25010}.
\end{defn}

@@

\begin{defn}
  We will view maintainability as two separate qualities: repairability and
  evolvability. Software is repairable if it allows the fixing of defects; it is
  evolvable if it allows changes that enable it to satisfy new requirements
  \citep{ghezzi1991fundamentals}.
\end{defn}

@@

\begin{defn}
  Effort required to locate and fix an error in an operational program
  \citep{McCallEtAl1977}. \wss{It looks
    like \citet{pressman2005software} was using \citet{McCallEtAl1977} for their
    definition.} \wss{PM - can you please look into whether
    \citet{pressman2005software} cite \citet{McCallEtAl1977}?  Do they give a
    reason for dropping the word operational?}  \pmi{Yes,
    \citet{pressman2005software} does cite the relevant section of the textbook
    with \citet{McCallEtAl1977}. While they do not give a reason for
    specifically dropping the word operational, they do indicate that their
    given definition ``is a very limited definition''. Perhaps their intent was
    breadth.} \wss{Peter, please verify the McCall reference definition, and
    then just cite that.  Olu has found the McCall reference, so you shouldn't
    have to search for it.  We can remove both the VanVliet and Pressman
    reference in  preference of the McCall definition.}
\pmi{I have verified the McCall definition and removed the Pressman and van Vliet references above. I will leave these comments until all similar definitions in this document have been edited.}
\end{defn}

\begin{defn}
  The capability of the software to be modified.  Modifications may include
  corrections, improvements or adaptation of the software to changes in
  environment, and in requirements and functional specifications. \cite{ISO9126}
\end{defn}

@@

\wss{What are the other attributes?  What is an attribute?  Is it the
	same as a quality?  We are calling performance a quality in this document.}
\pmi{The term quality attribute has come up often in the literature - qualities
	appear to be a subset of attributes. Quality attributes seem to be associated
	with non-functional requirements. The term resource attribute was encountered in
	a journal and implied an association with a functional requirement. The two
	might differ along those lines of NFR and FR.}

@@

\subsection{{Reusability} \pmi{owner}}

\begin{defn} 
  The degree to which a software module or other work product can be used in
  more than one software system \citep{IEEEStdGlossarySET1990}.
\end{defn}
\begin{defn}
  Extent to which a program can be used in other applications - related to the packaging and scope of the functions that programs perform \citep{McCallEtAl1977}. 
\end{defn}
\begin{defn} \label{ReusabilityDefnSelected} 
  The extent to which a software component can be used with or without
  adaptation in a problem solution other than the one for which it was
  originally developed \citep{kalagiakos2003non}.
\end{defn}
\begin{defn}
  Reusability is the likelihood a segment of source code that can be used again
  to add new functionalities with slight or no modification
  \citep{sandhu2010survey}.
\end{defn}

\begin{defn}
  A product is reusable if we can ``use the product - perhaps with minor changes
  - to build another product'' \citep{ghezzi1991fundamentals}.
\end{defn}

@@

@@

\begin{defn}
  Portability refers to the ability to run a system on different hardware
  platforms \citep{ghezzi1991fundamentals}.
\end{defn}

\begin{defn}
  The capability of software to be transferred from one environment to another
  \cite{ISO9126}.
\end{defn}
\noindent \textbf{Proposed Definition}

@@

@@

Interestingly, the business literature seems to have taken more care to define
this.  Here we encounter
\begin{quote}
  Understandability is the concept that X should be presented so that a reader
  can easily comprehend it.
\end{quote}
At least this brings in the idea that the \emph{reader} is actively involved,
and indirectly that the reader's knowledge may be relevant, as well as the
``clarity of exposition'' of X.

@@

\begin{defn}
  The capability of the software product to enable the user to understand
  whether the software is suitable, and how it can be used for particular tasks
  and conditions of use \cite{ISO9126}.
\end{defn}

@@

\begin{defn}
  \label{InteroperabilitySelected}
  The degree to which two or more systems, products or components can exchange
  information and use the information that has been exchanged
  \citep{ISO/IEC25010}.
\end{defn}

\begin{defn}
  The ability of a system to coexist and cooperate with other systems.
  \citep{ghezzi1991fundamentals}.
\end{defn}

\begin{defn}
  The capability to communicate, execute programs, and transfer data among
  various functional units in a manner that requires the user to have little or
  no knowledge of the unique characteristics of those units
  \citep{ISO/IEC/IEEE24765}.
\end{defn}

@@

\begin{defn}
  The capability of the software to interact with one or more specified systems
  \cite{ISO9126}.
\end{defn}
\noindent \textbf{Proposed Definition}

@@

\begin{defn}
  \label{VisibilitySelected}
  A software development process is visible if all of its steps and its current
  status are documented clearly. Another term used to characterize this property
  is transparency \citep{ghezzi1991fundamentals}.
\end{defn}

@@

The importance of reproducibility is seen by retractions of influential papers,
like the review of
\href{https://www.theguardian.com/commentisfree/2020/jun/05/lancet-had-to-do-one-of-the-biggest-retractions-in-modern-history-how-could-this-happen}
{hydroxychloroquine trials}

@@

\begin{defn}
  A quality of the software production process, referring to its efficiency and
  performance \citep{ghezzi1991fundamentals}.
\end{defn}
\begin{defn}
  The best definition of the productivity of a process is
	\[\text{Productivity} = \dfrac{\text{Outputs produced by the
              process}}{\text{Inputs consumed by the process}}\]

@@

\textbf{Defining outputs.} The big problem in defining software productivity is
defining outputs. Here we find a defining delivered source instructions (DSI) or
lines of code as the output of the software process is totally inadequate, and
they argue that there are a number of deficiencies in using DSI. However,
	most organizations doing practical productivity measurement still use DSI as
	their primary metric \citep{Boehm1987}. \wss{Is this a direct quote from
		\citet{Boehm1987}?  The sentences seem incomplete?}
\ad{I added some deleted parts, now it is a direct quote. After the ending of
the quote, the following 2 pages discuss the flaws of DCI and a list of
alternatives to DCI, so I ended the quote here.}
\end{defn}
\begin{defn}
	\label{ProductivitySelected1}
Productivity is the amount of output (what is produced) per unit of input used.
In general, productivity is difficult to measure because outputs and inputs are
typically quite diverse and are often themselves difficult to measure. In the
context of software, productivity measurement is usually based on a simple ratio
of product size to project effort. Thus,If we can measure the size of the
software product and the effort required to develop the product, we have:
	\begin{align}
	\text{productivity} = \text{size}/\text{effort}
	\end{align}
Equation (1) assumes that size is the output of the software production process
and effort is the input to the process. This can be contrasted with the
	viewpoint of software cost models where we use size as an independent variable
	(i.e., an input) to predict effort which is treated as an output. Equation (1)
	is simple to operationalize if we have a single dominant size measure, for
	example, product size measured in lines of code \citep{Kitchenham2004}.
\end{defn}
\begin{defn}
	\label{ProductivitySelected2}
The number of lines of new code developed per person-day (an imperfect measure
of productivity but one that could be measured consistently)
\citep{MacCormack2003}.
\end{defn}

@@

\begin{description}
\item[Intrinsic] Pertaining to characteristics of the software
\item[Extrinsic] Pertaining to the software development environment
\end{description}
Find paper that combines nonfunctional qualities into sustainability.

@@

To achieve the qualities listed in Section~\ref{SecQualities}, the documentation
should achieve the qualities listed in this section.  All but the final quality
listed (abstraction), are adapted from the IEEE recommended practise for
producing good software requirements \citep{IEEE1998}.  Abstraction means only
revealing relevant details, which in a requirements document means stating what
is to be achieved, but remaining silent on how it is to be achieved.
Abstraction is an important software development principle for dealing with
complexity \citep[p.~40]{GhezziEtAl2003}.  Correctness was in the above list, so
it is not repeated here.  \citet{SmithAndKoothoor2016} present further details
on the qualities of documentation for SCS.

>>>>

\usepackage{natbib}
\usepackage{booktabs}
\usepackage{amsmath,amsthm}
\usepackage{hyperref}
\hypersetup{
colorlinks=true,        % false: boxed links; true: colored links
linkcolor=red,          % color of internal links (change box color with linkbordercolor)
citecolor=blue,         % color of links to bibliography
filecolor=magenta,      % color of file links
urlcolor=cyan           % color of external links
}

@@

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007</concept_id>
<concept_desc>Software and its engineering</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering}

\keywords{software qualities, installability, correctness, verfiability,
  validatability, testability, reliability, robustness, performance,
  usability, maintainability, reusability, portability, understandability,
  interoperability, visibility/transparency, reproducibility, productivity,
  sustainability, completness, consistency, modifiability, traceability,
  unambiguity, verifiability, abstract}

@@

\begin{abstract}
	Existing definitions of software qualities (also called quality
  characteristics) are collected and analyzed.  For each quality a
  ``best'\rq{} definition is recommended based on the following criteria:
  unambiguity, simplicity, consistency, ...
\end{abstract}

@@

% Purpose

Given that our research focuses on improving the quality of software, we
often find it necessary to have clear definitions of the different software
qualities, like maintainability, usability, sustainability, etc.
Unfortunately, the definitions that we find in the literature are not always
satisfying.  This became clear when we recently undertook an exercise for
measuring the \lq\lq{}state of the practice\rq\rq{} in different software
domains, including the domains of seismology software \citep{SmithEtAl2018},
Geographic Information Systems \citep{SmithEtAl2018_arXivGIS} and mesh
generators \citep{smith2016state}. In trying to determine methods to quickly
assess quality of large samples of existing software, we found communication
between ourselves and other challenging.  We craved unambigous and
standardized definitions to enable methods for measuring qualities.
Therefore, we have endeavored to collect an exhaustive list of definitions
and from this build a set of recommended definitions, based on our quality
criteria for high-quality quality definitions.  We are publishing our
collection, since others may likewise find them valuable.

% Scope

Not specifically emphasizing scientific computing, but when a specific
context is needed for a definition, the scientific context will be used.  For
instance, the ISO definition of accuracy is defined in terms of precision, but
in scientific computing accuracy and precision are two different concepts.

\wss{List of definitions don't include textbooks that simply repeat
definitions from the literature.}

% Methodology
Methodology:

  \item Collect the list of qualities to define, based on the qualities that are
  listed in software engineering texts, such as Ghezzi Et al and Van Hliet.

  \item For each quality find as many definitions as possible

  \item Based on the existing definitions, propose a high quality definition of
  each quality.  The proposed definition may use one of the existing, or combine
  several of the options.  We may also make other modifications.  The goal is
  for the proposed definition to address our acceptance criteria.

\end{itemize}

@@

\wss{introduce the terminology of quality characteristics, as in ISO9126?}

\wss{tell the reader that the definitions will be sorted in the order of their
publication.}

@@

\begin{description}

  \item[Unambiguous]

  \item[Consistency] The final list of definitions should be consistent in
  the terminology used.  For instance, we do not use synonyms for variety;
  we use the same word to mean the same concept throughout the final list of
  definitions.  This is why measurement is given by ``the effort required.''
  and not the various synonyms for this that are seen in the literature.
  \wss{We may change this phrase we use for measuring something.  I just
  wanted to put this in here as a placeholder.}

  \item[Generality]

  \item[Measurability]

  \item[Etc.]

\end{description}

@@

\wss{add a definition related to capability?  The word seems to come up in  many definitions.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Installability \oo{owner}}

\begin{defn}
  The capability of the software product to be installed in a specified
  environment \citep{ISO9126}. %2001
\end{defn}

\begin{defn} \label{Installability_Selected2}
	The degree of effectiveness and efficiency with which a product or system can
	be successfully installed and/or uninstalled in a specified environment
	\citep{ISO/IEC25010}. %2011
\end{defn}

@@

\wss{Remove the repetition of the word being defined in the definition?  Remove
  reinstallation?  This quality characteristic is a product quality.  It is an
  external quality.  Installability should be something that can be measured
  directly.}

@@

 \wss{remove re-installation - it doesn't seem different from installation.
 Connection between installability and portability.  Use the terminology effort
 (without cost) to be consistent}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

 \subsection{Correctness \oo{owner}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{defn}
	Verifiability refers to how well software components or the integrated
  product can be evaluated to demonstrate whether the system functions as
  expected \citep{wiegers2003softreq}.
\end{defn}

\begin{defn} \label{Defn_Verifiability1}
  Verifiability ``means that you should be able to write a set of tests that
  can demonstrate that the delivered system meets each specified
  requirement'' \citep{sommerville}. %2011
\end{defn}

@@

\noindent \textbf{Proposed Definition} Definition~\ref{Defn_Verifiability1}
rephrased: Verifiability is the extent to which a set of tests can be
written and executed, to demonstrate that the delivered system meets the
specification.

@@

Definition is concise and measurable.  Verifiability involves solving the
equations right \citep[p.~23]{Roache1998} and the definition states that
system needs to be verifiable based on the specifications.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{defn}
  Software validation is achieved through a series of tests that demonstrate
  conformity with requirements \citep{pressman2005software}.
\end{defn}

@@

\begin{defn}
  Software validation is the process of checking that the system conforms to
  its specification and that it meets the real needs of the users of the
  system \citep{sommerville}. %2011
\end{defn}

\begin{defn}
	Software validation, where the software is checked to ensure that it is
  what the customer requires \citep{sommerville}. %2011
\end{defn}

\begin{defn}
  The assurance that a product, service, or result meets the needs of the
  customer and other identified stakeholders \citep{project2017guide}.
\end{defn}

@@

Definition describes that system should be measurable through the
development life cycle, It is concise and can be measured.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{\notdone{Testability} \oo{owner}}

@@

Definition is concise and measurable.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{defn} \label{reliabilitySelected1} It is the probability of failure-free
  operation of a computer program in a specified environment for a specified
  time \citep{musa1987software}.
\end{defn}

\begin{defn}
 The capability of the software product to maintain a specified level of
 performance when used under specified conditions \cite{ISO9126}. %2001
\end{defn}

@@

\begin{defn} \label{reliabilitySelected2} Reliability is concerned with measuring
	the probability of occurence of faliure - that is, the probability of the
	observable effects of errors \citep{GhezziEtAl2003}.
\end{defn}

\begin{defn}
  The ability of a system to perform its required functions under stated
  conditions whenever required—having a long mean time between failures
  \citep{mcconnell2004code}.
\end{defn}

\begin{defn}
  Informally, the reliability of a system is the probability, over a given
  period of time, that the system will correctly deliver services as expected by
  the user \citep{sommerville}. %2011
\end{defn}

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Robustness \pmi{owner}}

\begin{defn}
	The degree to which a system or component can function correctly in the
	presence of invalid inputs or stressful environmental conditions
	\citep{IEEEStdGlossarySET1990}.
\end{defn}

\begin{defn} \label{RobustnessDefnSelected}
	A program is robust if it behaves ``reasonably'', even in circumstances that
	were not anticipated in the requirements specification - for example, when it
	encounters incorrect input data or some hardware malfunction
	\citep{ghezzi1991fundamentals}. \wss{Use Ghezzi 2003 instead?}
\end{defn}

\begin{defn}
	The quality can be further informally refined as the ability of a software
  to keep an acceptable behaviour, expressed in terms of robustness
  requirements, in spite of exceptional or unforeseen execution conditions
  (such as the unavailability of system resources, communication failures,
  invalid or stressful inputs, etc.) \citep{fernandez2005model}.
\end{defn}

\begin{defn} \label{RobustnessPartialDefn}
	Code possesses the characteristic of robustness to the extent that it can
	continue to perform despite some violation of the assumptions in its
	specification \citep{boehm2007software}.
\end{defn}

@@

\begin{defn} \label{PerformanceDefnSelected}
  The degree to which a system or component accomplishes its designated
  functions within given constraints, such as speed, accuracy, or memory
  usage \citep{IEEEStdGlossarySET1990}.
\end{defn}

\begin{defn}
	In software engineering we often equate performance with efficiency. A
  software system is efficient if it uses computing resources economically
  \citep{ghezzi1991fundamentals}.  \wss{Change to 2003 reference?}
\end{defn}

\begin{defn} \label{PerformanceDefnSelected2}
	How well or how rapidly the system must perform specific functions. Performance
	requirements encompass speed (database response times, for instance), throughput
	(transactions per second), capacity (concurrent usage loads), and timing (hard
	real-time demands) \citep{wiegers2003softreq}.
\end{defn}

@@

\pmi{I will check if Wiegers discusses the difference between speed and
timing.}

\pmi{Wiegers did not discuss this difference.}

\pmi{Speed - change over time; timing - related to specification: when
something should happen in relation to something else - state transitions}

\wss{If we need both speed and timing, we should define them.  We aren't
  including accuracy from the IEEE definition.  Is that on purpose?}

\pmi{We have decided not to include accuracy. SS will find difference b/w
speed and timing from a colleague.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{defn}
  ``A software is usable - or user friendly - if its human users find it easy
  to use'' \citep{ghezzi1991fundamentals}. \wss{switch to 2003 ref?}
\end{defn}

@@

Nielsen and (separately) Schneidermann have defined usability as part of
usefulness and is composed of:
\begin{defn}
\begin{itemize}
\item Learnability: How easy is it for users to accomplish basic tasks the first
  time they encounter the design?
\item Efficiency: Once users have learned the design, how quickly can they
  perform tasks?
\item Memorability: When users return to the design after a period of not using
  it, how easily can they re-establish proficiency?
\item Errors: How many errors do users make, how severe are these errors, and
  how easily can they recover from the errors?
\item Satisfaction: How pleasant is it to use the design?
\end{itemize}
In that context, it makes sense to separate \emph{usefulness} into
\emph{usability} (purely an interface concern) and \emph{utility} (in the
economics sense of the word).
\cite{NielsenNorman} %2012
\end{defn}

@@

@@

\begin{defn}
  Effort required to locate and fix an error in an operational program
  \citep{McCallEtAl1977}.

  \wss{It looks like \citet{pressman2005software} was using
  \citet{McCallEtAl1977} for their definition.} \wss{PM - can you please
  look into whether \citet{pressman2005software} cite
  \citet{McCallEtAl1977}?  Do they give a reason for dropping the word
  operational?}  \pmi{Yes, \citet{pressman2005software} does cite the
  relevant section of the textbook with \citet{McCallEtAl1977}. While they
  do not give a reason for specifically dropping the word operational, they
  do indicate that their given definition ``is a very limited definition''.
  Perhaps their intent was breadth.} \wss{Peter, please verify the McCall
  reference definition, and then just cite that.  Olu has found the McCall
  reference, so you shouldn't have to search for it.  We can remove both the
  VanVliet and Pressman reference in preference of the McCall definition.}

  \pmi{I have verified the McCall definition and removed the Pressman and
  van Vliet references above. I will leave these comments until all similar
  definitions in this document have been edited.}
\end{defn}

@@

\begin{defn}
  We will view maintainability as two separate qualities: repairability and
  evolvability. Software is repairable if it allows the fixing of defects;
  it is evolvable if it allows changes that enable it to satisfy new
  requirements \citep{ghezzi1991fundamentals}. \wss{update to 2003 ref?}
\end{defn}

\begin{defn}
  The capability of the software to be modified.  Modifications may include
  corrections, improvements or adaptation of the software to changes in
  environment, and in requirements and functional specifications.
  \cite{ISO9126} %2001
\end{defn}

@@

@@

\begin{defn}
  ISO/IEC 25010 refers to maintainability as the degree of effectiveness and
  efficiency with which a product or system can be modified by the intended
  maintainers \citep{ISO/IEC25010}. %2011
\end{defn}

@@

\wss{What are the other attributes?  What is an attribute?  Is it the same
as a quality?  We are calling performance a quality in this document.}

\pmi{The term quality attribute has come up often in the literature -
qualities appear to be a subset of attributes. Quality attributes seem to be
associated with non-functional requirements. The term resource attribute was
encountered in a journal and implied an association with a functional
requirement. The two might differ along those lines of NFR and FR.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Reusability \pmi{owner}}

\begin{defn}
  Extent to which a program can be used in other applications - related to
  the packaging and scope of the functions that programs perform
  \citep{McCallEtAl1977}.
\end{defn}

\begin{defn}
  The degree to which a software module or other work product can be used in
  more than one software system \citep{IEEEStdGlossarySET1990}.
\end{defn}

\begin{defn}
  A product is reusable if we can ``use the product - perhaps with minor changes
  - to build another product'' \citep{ghezzi1991fundamentals}.  \wss{Update
  to 2003?}
\end{defn}

\begin{defn} \label{ReusabilityDefnSelected}
  The extent to which a software component can be used with or without
  adaptation in a problem solution other than the one for which it was
  originally developed \citep{kalagiakos2003non}.
\end{defn}

\begin{defn}
  Reusability is the likelihood a segment of source code that can be used
  again to add new functionalities with slight or no modification
  \citep{sandhu2010survey}.
\end{defn}

@@

\begin{defn}
  Portability refers to the ability to run a system on different hardware
  platforms \citep{ghezzi1991fundamentals}.  \wss{update to 2003?}
\end{defn}

\begin{defn}
  The capability of software to be transferred from one environment to
  another \cite{ISO9126}. %2001
\end{defn}

@@

@@

\begin{defn}
  The capability of the software product to enable the user to understand
  whether the software is suitable, and how it can be used for particular tasks
  and conditions of use \cite{ISO9126}. %2001
\end{defn}

@@

Interestingly, the business literature seems to have taken more care to
define this.  Here we encounter:

\begin{quote}
  Understandability is the concept that X should be presented so that a reader
  can easily comprehend it.
\end{quote}

At least this brings in the idea that the \emph{reader} is actively involved,
and indirectly that the reader's knowledge may be relevant, as well as the
``clarity of exposition'' of X.

@@

@@

\begin{defn}
  The ability of a system to coexist and cooperate with other systems.
  \citep{ghezzi1991fundamentals}.  \wss{2003 ref?}
\end{defn}

\begin{defn}
  The capability of the software to interact with one or more specified
  systems \cite{ISO9126}. %2001
\end{defn}

\begin{defn}
  The capability to communicate, execute programs, and transfer data among
  various functional units in a manner that requires the user to have little
  or no knowledge of the unique characteristics of those units
  \citep{ISO/IEC/IEEE24765}. %2010
\end{defn}

\begin{defn}
  \label{InteroperabilitySelected} The degree to which two or more systems,
  products or components can exchange information and use the information
  that has been exchanged \citep{ISO/IEC25010}. %2011
\end{defn}

@@

@@

\begin{defn}
  \label{VisibilitySelected} A software development process is visible if
  all of its steps and its current status are documented clearly. Another
  term used to characterize this property is transparency
  \citep{ghezzi1991fundamentals}. \wss{change to 2003 ref?}
\end{defn}

@@

The importance of reproducibility is seen by retractions of influential
papers, like the review of \href{https://www.theguardian
.com/commentisfree/2020/jun/05/lancet-had-to-do-one-of-the-biggest
-retractions-in-modern-history-how-could-this-happen} {hydroxychloroquine
trials}

@@

\begin{defn}
  The best definition of the productivity of a process is
	\[\text{Productivity} = \dfrac{\text{Outputs produced by the
              process}}{\text{Inputs consumed by the process}}\]

@@

\textbf{Defining outputs.} The big problem in defining software productivity is
defining outputs. Here we find a defining delivered source instructions (DSI) or
lines of code as the output of the software process is totally inadequate, and
they argue that there are a number of deficiencies in using DSI. However,
	most organizations doing practical productivity measurement still use DSI as
	their primary metric \citep{Boehm1987}. \wss{Is this a direct quote from
		\citet{Boehm1987}?  The sentences seem incomplete?}
\ad{I added some deleted parts, now it is a direct quote. After the ending of
the quote, the following 2 pages discuss the flaws of DCI and a list of
alternatives to DCI, so I ended the quote here.}
\end{defn}

\begin{defn}
  A quality of the software production process, referring to its efficiency
  and performance \citep{ghezzi1991fundamentals}.  \wss{2003 ref?}
\end{defn}

\begin{defn}
	\label{ProductivitySelected2} The number of lines of new code developed per
  person-day (an imperfect measure of productivity but one that could be
  measured consistently) \citep{MacCormack2003}.
\end{defn}

\begin{defn}
	\label{ProductivitySelected1} Productivity is the amount of output (what is
  produced) per unit of input used. In general, productivity is difficult to
  measure because outputs and inputs are typically quite diverse and are
  often themselves difficult to measure. In the context of software,
  productivity measurement is usually based on a simple ratio of product
  size to project effort. Thus,If we can measure the size of the software
  product and the effort required to develop the product, we have:
	\begin{align}
	\text{productivity} = \text{size}/\text{effort}
	\end{align}
  Equation (1) assumes that size is the output of the software production
  process and effort is the input to the process. This can be contrasted
  with the viewpoint of software cost models where we use size as an
  independent variable (i.e., an input) to predict effort which is treated
  as an output. Equation (1) is simple to operationalize if we have a single
  dominant size measure, for example, product size measured in lines of code
  \citep{Kitchenham2004}.
\end{defn}

@@

\begin{description}
\item[Intrinsic] Pertaining to characteristics of the software
\item[Extrinsic] Pertaining to the software development environment
\end{description}

Find paper that combines nonfunctional qualities into sustainability.

@@

To achieve the qualities listed in Section~\ref{SecQualities}, the
documentation should achieve the qualities listed in this section.  All but
the final quality listed (abstraction), are adapted from the IEEE
recommended practise for producing good software requirements
\citep{IEEE1998}.  Abstraction means only revealing relevant details, which
in a requirements document means stating what is to be achieved, but
remaining silent on how it is to be achieved. Abstraction is an important
software development principle for dealing with complexity \citep[p
.~40]{GhezziEtAl2003}.  Correctness was in the above list, so it is not
repeated here.  \citet{SmithAndKoothoor2016} present further details on the
qualities of documentation for SCS.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
